Adaptive Exponentiated Gradient

### 2014-01-27 lp/introduction.tex
mostly fixed equation formatting

By exploiting this connection, we show that the perhaps more natural update 
of $\exp(-\eta z_{t,i} - \eta^2 z_{t,i}^2)$ achieves the same regret bound 
as in (\ref{eqn:mw-regret}), and that both updates continue to enjoy this 
regret even when generalized from the simplex to an arbitrary convex set 
(including all of $\bR^n$, which corresponds to unconstrained exponentiated 
gradient).

@@

For convenience, we summarize our contributions below:
\begin{itemize}
\item An interpretation of the multiplicative weights update as 
      exponentiated gradient with an adaptive regularizer.
\item An improved exponentiated gradient algorithm based on adaptive 
      updates to an entropic regularizer.
\item Further improvements to exponentiated gradient based on optimistic 
      predictions and online learning of the mean gradient.
\item An adaptive matrix exponentiated gradient, with similar improvements 
      based on learning the mean gradient.
\item A generalization of Follow the Regularized Leader to vector-valued 
      loss functions.
\end{itemize}

>>>>

% By exploiting this connection, we show that the perhaps more natural update 
% of $\exp(-\eta z_{t,i} - \eta^2 z_{t,i}^2)$ achieves the same regret bound 
% as in (\ref{eqn:mw-regret}), and that both updates continue to enjoy this 
% regret even when generalized from the simplex to an arbitrary convex set 
% (including all of $\bR^n$, which corresponds to unconstrained exponentiated 
% gradient).

@@

For convenience, we summarize our contributions below:
\begin{itemize}
\item An interpretation of the multiplicative weights update as 
      exponentiated gradient with an adaptive regularizer (Section~\ref{sec:mw12}).
%\item An improved exponentiated gradient algorithm based on adaptive 
%      updates to an entropic regularizer.
\item An improved exponentiated gradient algorithm based on an adaptive 
      regularizer, optimistic predictions, and online learning of the 
      mean gradient (Sections~\ref{sec:machinery} and~\ref{sec:ftrl-aux}).
\item An adaptive matrix exponentiated gradient algorithm, 
      with similar improvements based on learning the mean gradient 
      (Section~\ref{sec:matrix}).
\item A generalization of Follow the Regularized Leader to vector-valued 
      loss functions (Lemma~\ref{lem:ftrl-k}).
\end{itemize}

### 2014-01-27 lp/machinery.tex
fixed missing references
mostly fixed equation formatting

\begin{algorithm}
\caption{Adaptive Optimistic Mirror Descent}
\label{alg:adaptive-optimistic}
\begin{algorithmic}
\STATE $\theta_1 = 0$
\FOR{$t=1$ {\bfseries to} $T$}
  \STATE Choose $w_t = \nabla \psi_t^*(\theta_t - \eta m_t)$
  \STATE Observe $z_t$ and suffer loss $w_t^{\top}z_t$
  \STATE Update $\theta_{t+1} = \theta_t - \eta z_t$
\ENDFOR
\end{algorithmic}
\end{algorithm}

The key regret bound for Algorithm~\ref{alg:adaptive-optimistic} is 
given in Theorem~\ref{thm:adaptive-optimistic}:
\begin{theorem}
Algorithm~\ref{alg:adaptive-optimistic} achieves regret bounded by
\begin{equation}
\frac{1}{\eta}\left(\psi_1^*(\theta_1) + \psi_{T+1}(w) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t - \eta m_t) + \eta w_t^{\top}m_t\right).
\end{equation}
\end{theorem}
\begin{proof}
TODO
\end{proof}
As an immediate consequence, we have
\begin{corollary}
\label{cor:a-o1}
Suppose that $\psi_{t+1}$ is chosen so that $\psi_{t+1}^*(\theta_t - \eta z_t) \leq \psi_t^*(\theta_t - \eta m_t) - \eta w_t^{\top}(z_t - m_t)$. Then
\begin{equation}
\Regret(w) \leq \frac{\psi_1^*(\theta_1) + \psi_{T+1}(w)}{\eta}.
\end{equation}
\end{corollary}
\begin{proof}
TODO
\end{proof}
Finally, we consider a simplified version of Algorithm~\ref{alg:adaptive-optimistic}, 
which corresponds to choosing $\psi_t(w) = \psi(w) + \eta^2 w^{\top} \sum_{s=1}^{t-1} u_s$ 
for some fixed function $\psi$.

@@

\begin{algorithmic}
\STATE Initialize $\theta_1$ arbitrarily
\FOR{$t=1$ {\bfseries to} $T$}
  \STATE Choose $w_t = \nabla \psi^*(\theta_t - \eta m_t)$
  \STATE Observe $z_t$ and suffer loss $w_t^{\top}z_t$
  \STATE Update $\theta_{t+1} = \theta_t - \eta z_t - \eta^2 u_t$
\ENDFOR
\end{algorithmic}
\end{algorithm}

\begin{corollary}
\label{cor:a-o2}
Suppose that $u_t$ is chosen so that $\psi^*(\theta_t - \eta z_t - \eta^2 u_t) \leq \psi^*(\theta_t - \eta m_t) - \eta w_t^{\top}(z_t - m_t)$. 
Then 
\begin{equation}
\Regret(w) \leq \frac{\psi^*(\theta_1)}{\eta} + \eta w^{\top} \sum_{t=1}^T u_t.
\end{equation}
\end{corollary}
\begin{proof}
TODO
\end{proof}
To give some intuition for the condition in Corollary~\ref{cor:a-o2}, note that 
$w_t = \nabla \psi^*(z_t - \eta m_t)$, and so, since $\psi^*$ is convex, we have
$\psi^*(\theta_t - \eta z_t) \geq \psi^*(\theta_t - \eta m_t) - \eta w_t^{\top}(z_t - m_t)$. 
The term $\eta^2 u_t$ is thus a second-order correction to flip the sign of the inequality 
(by taking a Taylor expansion, one can show that $\eta^2$ is the right order of magnitude for 
the correction term).

\paragraph{Application to exponentiated gradient.} We now show how to apply 
Corollary~\ref{cor:a-o2} to obtain an adaptive exponentiated gradient algorithm.
\begin{proposition}[Adaptive Exponentiated Gradient]
\label{prop:aeg}
Consider the updates given by $\theta_{1,i} = -\log(n)$ and 
$\theta_{t+1,i} = \theta_{t,i} - \eta z_{t,i} - \eta^2 (z_{t,i} - m_{t,i})^2$, 
with prediction $w_{t,i} = \exp(\theta_{t,i} - \eta m_{t,i})$. For this 
algorithm,
\begin{equation}
\Regret(w) \leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2.
\end{equation}
\end{proposition}
\begin{proof}
TODO
\end{proof}
Note that in the case that $m_{t,i} = 0$, the update is 
$\theta_{t+1,i} = \theta_{t,i} - \eta z_{t,i} - \eta^2 z_{t,i}^2$, whereas multiplicative 
updates by $1 - \eta z_{t,i}$ would correspond to 
$\theta_{t+1,i} = \theta_{t,i} + \log(1 - \eta z_{t,i})$. Since 
$-\eta z_{t,i} - \eta^2 z_{t,i}^2 \leq \log(1 - \eta z_{t,i})$ when 
$|\eta z_{t,i}| \leq \frac{1}{2}$, we can think of these updates as a second-order 
under-approximation to (\ref{eqn:mw2}), which however can more naturally incorporate 
the optimistic updates based on $m_t$.

>>>>

\begin{algorithm}
\caption{Adaptive Optimistic Mirror Descent}
\label{alg:adaptive-optimistic}
\begin{algorithmic}
\STATE $\theta_1 = 0$
\FOR{$t=1$ {\bfseries to} $T$}
  \STATE Choose $w_t = \nabla \psi_t^*(\theta_t - \eta m_t)$
  \STATE Observe $z_t$ and suffer loss $w_t^{\top}z_t$
  \STATE Update $\theta_{t+1} = \theta_t - \eta z_t$
\ENDFOR
\end{algorithmic}
\end{algorithm}
Note that Algorithm~\ref{alg:adaptive-optimistic} does not contain a 
projection step back onto the feasible domain $S$. If $\psi_t$ is defined to be 
infinite outside of $S$, then $w_t$ will always lie in $S$ and hence the 
projection step is unnecessary. (However, modifying $\psi_t$ in this way may 
complicate the analysis, i.e. by giving $\psi_t^*$ a more complicated form.)

The regret bound for Algorithm~\ref{alg:adaptive-optimistic} is 
given in Theorem~\ref{thm:adaptive-optimistic}. A key observation is that 
the regret is in terms of $z_t - m_t$ rather than $z_t$ (so the closer 
$m_t$ is to $z_t$, the smaller the regret; this will be more apparent 
in various corollaries).
\begin{theorem}
\label{thm:adaptive-optimistic}
Algorithm~\ref{alg:adaptive-optimistic} achieves regret bounded by
\begin{dmath}
\label{eqn:regret-1}
\frac{1}{\eta}\left(\psi_1^*(\theta_1) + \psi_{T+1}(w) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t - \eta m_t) + \eta w_t^{\top}(z_t-m_t)\right).
\end{dmath}
\end{theorem}
\begin{proof}
First note that $\psi_t^*$ is convex and that 
$w_t = \nabla \psi_t^*(\theta_t - \eta m_t)$. It thus follows that 
$\psi_t^*(\theta_t) \geq \psi_t^*(\theta_t - \eta m_t) + \eta m_t^{\top}z_t$. 
Then, by definition of the Fenchel conjugate together with telescoping sums, we 
have, for any $w$, 
\begin{align*}
\lefteqn{\hskip -0.1in w^{\top}\theta_{T+1} - \psi_{T+1}(w)} \\
\hskip 0.1in  &\leq \psi_{T+1}^*(\theta_{T+1}) \\
  &= \psi_1^*(\theta_1) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t) \\
  &\leq \psi_1^*(\theta_1) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t-\eta m_t) - \eta w_t^{\top}m_t.
\end{align*}
Re-arranging and expanding $\theta_{T+1}$, we have
\begin{dmath}
-\eta w^{\top} \sum_{t=1}^T z_t \leq \psi_{T+1}(w) + \psi_1^*(\theta_1) + \sum_{t=1}^{T} \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t-\eta m_t) - \eta w_t^{\top}m_t.
\end{dmath}
Adding $\eta \sum_{t=1}^T w_t^{\top}z_t$ to both sides, we obtain
\begin{dmath}
\eta \Regret(w) \leq \psi_{T+1}(w) + \psi_1^*(\theta_1) + \sum_{t=1}^T \psi_{t+1}^*(\theta_{t+1}) - \psi_t^*(\theta_t) + \eta w_t^{\top}(z_t-m_t),
\end{dmath}
which yields the desired result.
\end{proof}
As an immediate consequence, we have
\begin{corollary}
\label{cor:a-o1}
Suppose that $\psi_{t+1}$ is chosen so that $\psi_{t+1}^*(\theta_t - \eta z_t) \leq \psi_t^*(\theta_t - \eta m_t) - \eta w_t^{\top}(z_t - m_t)$. Then
\begin{equation}
\Regret(w) \leq \frac{\psi_1^*(\theta_1) + \psi_{T+1}(w)}{\eta}.
\end{equation}
\end{corollary}
\begin{proof}
Since $\theta_{t+1} = \theta_t - \eta z_t$, if the condition in the corollary 
holds then the sum in the regret bound (\ref{eqn:regret-1}) is bounded above 
term-wise by zero. 
Hence we obtain $\Regret(w) \leq \frac{1}{\eta}\left(\psi_1^*(\theta_1) + \psi_{T+1}(w)\right)$, as was to be shown.
\end{proof}
Finally, we consider a simplified version of Algorithm~\ref{alg:adaptive-optimistic}, 
which corresponds to choosing $\psi_t(w) = \psi(w) + \eta^2 w^{\top} \sum_{s=1}^{t-1} u_s$ 
for some fixed function $\psi$. In this case, we define 
$\psi_S^*(w) \eqdef \sup_{w \in S} w^{\top}x - \psi(w)$, which 
will allow us to more easily handle restrictions on the domain $S$.

@@

\begin{algorithmic}
\STATE Initialize $\theta_1$ arbitrarily
\FOR{$t=1$ {\bfseries to} $T$}
  \STATE Choose $w_t = \nabla \psi_S^*(\theta_t - \eta m_t)$
  \STATE Observe $z_t$ and suffer loss $w_t^{\top}z_t$
  \STATE Update $\theta_{t+1} = \theta_t - \eta z_t - \eta^2 u_t$
\ENDFOR
\end{algorithmic}
\end{algorithm}

\begin{corollary}
\label{cor:a-o2}
Suppose that $u_t$ is chosen so that $\psi^*(\theta_t - \eta z_t - \eta^2 u_t) \leq \psi^*(\theta_t - \eta m_t) - \eta w_t^{\top}(z_t - m_t)$. 
Then 
\begin{equation}
\Regret(w) \leq \frac{\psi_S^*(\theta_1)}{\eta} + \eta w^{\top} \sum_{t=1}^T u_t.
\end{equation}
\end{corollary}
Note that, importantly, the conditions of Corollary~\ref{cor:a-o2} are in terms 
of $\psi^*$, while the regret bound and the choice of $w_t$ are in terms of 
$\psi_S^*$. This means that, once we construct an algorithm for the 
unconstrained case, we then achieve a regret bound under arbitrary convex 
constraints.
\begin{proof}
TODO
\end{proof}
To give some intuition for the condition in Corollary~\ref{cor:a-o2}, note that 
$w_t = \nabla \psi^*(z_t - \eta m_t)$, and so, since $\psi^*$ is convex, we have
$\psi^*(\theta_t - \eta z_t) \geq \psi^*(\theta_t - \eta m_t) - \eta w_t^{\top}(z_t - m_t)$. 
The term $\eta^2 u_t$ is thus a second-order correction to flip the sign of the inequality 
(by taking a Taylor expansion, one can show that $\eta^2$ is the right order of magnitude for 
the correction term).

\paragraph{Application to exponentiated gradient.} We now show how to apply 
Corollary~\ref{cor:a-o2} to obtain an adaptive exponentiated gradient algorithm. 
We will state it here in the case of the simplex, since the bounds in this 
case are particularly easy to interpret. Note, however, that our analysis 
goes via Corollary~\ref{cor:a-o2}, and hence applies to arbitrary convex 
constraints.
\begin{proposition}[Adaptive Exponentiated Gradient]
\label{prop:aeg}
Consider the updates given by $\theta_{1,i} = 0$ and 
$\theta_{t+1,i} = \theta_{t,i} - \eta z_{t,i} - \eta^2 (z_{t,i} - m_{t,i})^2$, 
with prediction $w_{t,i} \propto \exp(\theta_{t,i} - \eta m_{t,i})$. Then, 
assuming $\|z_t\|_{\infty} \leq 1$, $\|m_t\|_{\infty} \leq 1$ and 
$0 < \eta \leq \frac{1}{4}$, we have
\begin{equation}
\Regret(w) \leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n w_i \sum_{t=1}^T (z_{t,i} - m_{t,i})^2.
\end{equation}
\end{proposition}
\begin{proof}
We note that, for $S$ equal to the simplex $\Delta_n$ and 
$\psi(w) = \sum_{i=1}^n w_i\log(w_i)$, the $w_t$ is equal to 
$\nabla \psi_S^*(\theta_t - \eta m_t)$. It therefore suffices 
to check that the condition of Corollary~\ref{cor:a-o2} is satisfied with 
$u_{t,i} = (z_{t,i}-m_{t,i})^2$. 
Note that $\psi^*(\theta) = \sum_{i=1}^n \exp(\theta_i)$. We then have
\begin{dgroup*}
\begin{dmath*} \lefteqn{\psi^*(\theta_{t} - \eta z_t - \eta^2 u_t)} 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta z_{t,i} - \eta^2 (z_{t,i} - m_{t,i})^2) 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i})\exp(-\eta (z_{t,i} - m_{t,i}) - \eta^2 (z_{t,i} - m_{t,i})^2) 
\end{dmath*}\begin{dmath*}  \leq \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i})(1 - \eta (z_{t,i} - m_{t,i})) 
\end{dmath*}\begin{dmath*}  = \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i}) - \eta \sum_{i=1}^n \exp(\theta_{t,i} - \eta m_{t,i}) (z_{t,i} - m_{t,i}) 
\end{dmath*}\begin{dmath*}  = \psi^*(\theta_t - \eta m_t) - \eta \nabla \psi^*(\theta_t - \eta m_t)^{\top} (z_t - m_t).
\end{dmath*}
\end{dgroup*}
The one inequality we made use of was $\exp(-x-x^2) \leq 1-x$ for $|x| < \frac{1}{2}$.
This verifies the condition of Corollary~\ref{cor:a-o2}, yielding a regret bound 
of $\frac{\psi_S^*(0) + \psi(w)}{\eta} + \eta \sum_{i=1}^n w^{\top} u_t$. 
Finally, we note that $\psi_S^*(0) = \log(n)$, 
$\psi(w) = \sum_{i=1}^n w_i\log(w_i) \leq 0$, and 
$u_{t,i} = (z_{t,i} - m_{t,i})^2$, which completes the proof.
\end{proof}
Note that in the case that $m_{t,i} = 0$, the update is 
$\theta_{t+1,i} = \theta_{t,i} - \eta z_{t,i} - \eta^2 z_{t,i}^2$, whereas multiplicative 
updates by $1 - \eta z_{t,i}$ would correspond to 
$\theta_{t+1,i} = \theta_{t,i} + \log(1 - \eta z_{t,i})$. Since 
$-\eta z_{t,i} - \eta^2 z_{t,i}^2 \leq \log(1 - \eta z_{t,i})$ when 
$|\eta z_{t,i}| \leq \frac{1}{2}$, we can think of these updates as a second-order 
under-approximation to (\ref{eqn:mw2}), which however can more naturally incorporate 
the optimistic updates based on $m_t$.

### 2014-01-27 lp/matrix.tex
mostly fixed equation formatting

We start with an extension of Proposition~\ref{prop:aeg} to the matrix setting:
\begin{proposition}
\label{prop:maeg}
Consider the updates given by $\Theta_1 = -\log(n)I$ and 
$\Theta_{t+1} = \Theta_t - \eta Z_t - \eta^2 (Z_t - M_t)^2$, 
with prediction $W_t = \exp(\Theta_t - \eta M_t)$. For this algorithm,
\begin{equation}
\Regret(w) \leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n \Tr(W(Z_t-M_t)^2).
\end{equation}
\end{proposition}
\begin{proof}
TODO
\end{proof}
We would now, as in Section~\ref{sec:ftrl-aux}, like to learn $M_t$ adaptively 
to be competitive with the best fixed value $M^*$. However, it isn't even clear 
that such a best fixed value exists (since the optimal value may depend on $W$). 
Lemma~\ref{lem:global-opt} addresses this issue (proved in slightly greater 
generality for later convenience):
\begin{lemma}
\label{lem:global-opt}
For any $\delta \geq 0$, define $M^* \eqdef \frac{1}{T+\delta} \sum_{t=1}^T Z_t$. Then, for 
any $M' \neq M^*$, we have
\[ \delta (M^*)^2 + \sum_{t=1}^T (Z_t-M^*)^2 \preceq \delta (M')^2 + \sum_{t=1}^T (Z_t-M')^2. \]
\end{lemma}
\begin{proof}
Write $M' = M^* + D$. Then we have
\begin{align}
\lefteqn{\delta(M')^2 + \sum_{t=1}^T (Z_t - M')^2} \\
 &= \delta(M^*+D)^2 + \sum_{t=1}^T (Z_t - M^* - D)^2 \\
 &= \delta(M^*)^2 + \sum_{t=1}^T (Z_t - M^*)^2 \\
\nonumber
 &\phantom{==} + \left[\delta M^* + \sum_{t=1}^T M^*-Z_t\right]D \\
\nonumber
 &\phantom{==} + D\left[\delta M^* + \sum_{t=1}^T M^*-Z_t\right] + TD^2 \\
 &= \delta(M^*)^2 + \sum_{t=1}^T (Z_t-M^*)^2 + TD^2 \\
 &\succeq \delta(M^*)^2 + \sum_{t=1}^T (Z_t-M^*)^2,
\end{align}
which completes the lemma.
\end{proof}
Setting $\delta$ to $0$, we now see that $\frac{1}{T} \sum_{t=1}^T Z_t$ is the 
optimal value of $M^*$ for any $W \succeq 0$, in analogy with the vector case. We 
remark that Lemma~\ref{lem:global-opt} is almost purely algebraic, and only relies 
on the property that $D^2 \succeq 0$ for any symmetric matrix $D$. We could obtain 
a similar lemma for any partially ordered algebra equipped with an adjoint operator 
$*$ such that $a^{*}a \succeq 0$ for all elements $a$ of the algebra.

@@

\begin{lemma}
\label{lem:ftrl-k}
Suppose that for all $1 \leq t \leq T+1$, there 
exists a global minimizer $M_t$ of 
$\psi(X) + \sum_{s=1}^{t-1} f_s(X)$. Then for all $M$,
\[ \sum_{t=1}^T f_t(M_t)-f_t(M) \leq_{\sK} \psi(M) - \psi(M_1) + \sum_{t=1}^T f_t(M_{t}) - f_t(M_{t+1}). \]
\end{lemma}
\begin{proof}
We will prove the lemma by induction on $T$. Note that the lemma is equivalent to showing that 
\[ \psi(M_1) + \sum_{t=1}^T f_t(M_{t+1}) \leq_{\sK} \psi(M) + \sum_{t=1}^T f_t(M) \]
for all $M$.

In the base case $T = 0$, we have
\[ \psi(M_1) \leq_{\sK} \psi(M), \]
which follows from the fact that $M_1$ is a global minimizer of $\psi$ and 
hence $\psi(M_1) \leq_{\sK} \psi(M)$ for all $M$. For the inductive step, suppose that
\[ \sum_{t=1}^{T-1} f_t(M_{t+1}) \leq_{\sK} \psi(M) \sum_{t=1}^{T-1} f_t(M) \]
for all $M$, and invoke this for the particular choice $M = M_{T+1}$. Then we have
\begin{align}
\lefteqn{\psi(M_1) + \sum_{t=1}^T f_t(M_{t+1})} \\
 &= \psi(M_1) + \sum_{t=1}^{T-1} f_t(M_{t+1}) + f_T(M_{T+1}) \\
 &\leq_{\sK} \psi(M_{T+1}) + \sum_{t=1}^{T-1} f_t(M_{T+1}) + f_T(M_{T+1}) \\
 &= \psi(M_{T+1}) + \sum_{t=1}^T f_t(M_{T+1}) \\
 &\leq_{\sK} \psi(M) + \sum_{t=1}^T f_t(M)
\end{align}
for all $M$, where we use the fact that $M_{T+1}$ is a global minimizer of $\psi(M) + \sum_{t=1}^T f_t(M)$ for 
the last inequality. This completes the inductive hypothesis and hence the proof.
\end{proof}

As a corollary, we have (compare to Proposition~\ref{prop:ftrl-aux}):
\begin{corollary}
\label{cor:ftrl-k-aux}
Suppose that we choose $M_t = \frac{1}{t} \sum_{s=1}^{t-1} Z_s$. Then
\[ \sum_{t=1}^T (Z_t - M_t)^2 \preceq \sum_{t=1}^T (Z_t - M^*)^2 + 4\log(T+1)I, \]
for $M^* \eqdef \frac{1}{T} \sum_{t=1}^T Z_t$.
\end{corollary}
Combining Proposition~\ref{prop:maeg} with Corollary~\ref{cor:ftrl-k-aux} gives a 
regret bound for the MEG-Variance algorithm, whose updates are given in 
Table~\ref{tab:algorithms}.
\begin{corollary}
\label{cor:maeg-variance}
The MEG-Variance algorithm achieves a regret bound of
\[ \Regret(w) \leq \frac{\log(n)}{\eta} + \eta \left[ \sum_{t=1}^T \Tr(W(Z_t-M^*)^2) + 4\log(T+1) \right]. \]
\end{corollary}

>>>>

We start with an extension of Proposition~\ref{prop:aeg} to the matrix setting:
\begin{proposition}
\label{prop:maeg}
Consider the updates given by $\Theta_1 = -\log(n)I$ and 
$\Theta_{t+1} = \Theta_t - \eta Z_t - \eta^2 (Z_t - M_t)^2$, 
with prediction $W_t = \frac{\exp(\Theta_t - \eta M_t)}{\Tr(\exp(\Theta_t - \eta M_t))}$. For this algorithm,
\begin{equation}
\Regret(w) \leq \frac{\log(n)}{\eta} + \eta \sum_{i=1}^n \Tr(W(Z_t-M_t)^2).
\end{equation}
\end{proposition}
\begin{proof}
When $\psi(W) = \Tr(W\log(W))$ and $S$ consists of positive semidefinite 
matrices with trace $1$, the prediction $W_t$ corresponds to 
$\nabla \psi_S^*(\Theta_t - \eta M_t)$. So, again, it suffices to check 
that the condition of Corolarry~\ref{cor:a-o2} is satisfies for 
$U_t = (Z_t - M_t)^2$. To do so, we need to make use of the 
Golden-Thompson inequality $\Tr(\exp(A+B)) \preceq \Tr(\exp(A)\exp(B))$, 
together with the fact that $-X-X^2 \preceq \log(I-X)$ for 
$-\frac{1}{2}I \preceq X \preceq \frac{1}{2}I$. We then have
\begin{align*}
\lefteqn{\psi^*(\Theta_t - \eta Z_t - \eta^2 U_t)} \\
 &= \Tr(\exp(\Theta_t - \eta Z_t - \eta^2 (Z_t - M_t)^2)) \\
 &\leq \Tr(\exp(\Theta_t - \eta M_t)\exp(-\eta (Z_t - M_t) - \eta^2 (Z_t - M_t)^2)) \\
 &\leq \Tr(\exp(\Theta_t - \eta M_t)(I - \eta (Z_t - M_t))) \\
 &= \Tr(\exp(\Theta_t - \eta M_t)) - \eta \Tr(\exp(\Theta_t - \eta M_t)(Z_t - M_t)) \\
 &= \psi^*(\Theta_t - \eta M_t) - \eta \langle \nabla \psi^*(\Theta_t - \eta M_t), Z_t - M_t \rangle.
\end{align*}
This verifies the condition of Corolarry~\ref{cor:a-o2}, so that we have 
a regret bound of $\frac{\psi_S^*(0) + \psi(W)}{\eta} + \eta \sum_{t=1}^T \Tr(WU_t)$. 
Finally, noting that $\psi_S^*(0) = \log(n)$, $\psi(W) = \Tr(W\log(W)) \leq 0$, and 
$U_t = (Z_t - M_t)^2$ completes the proof.
\end{proof}
We would now, as in Section~\ref{sec:ftrl-aux}, like to learn $M_t$ adaptively 
to be competitive with the best fixed value $M^*$. However, it isn't even clear 
that such a best fixed value exists (since the optimal value may depend on $W$). 
Lemma~\ref{lem:global-opt} addresses this issue (proved in slightly greater 
generality for later convenience):
\begin{lemma}
\label{lem:global-opt}
For any $\delta \geq 0$, define $M^* \eqdef \frac{1}{T+\delta} \sum_{t=1}^T Z_t$. Then, for 
any $M' \neq M^*$, we have
\begin{equation*} \delta (M^*)^2 + \sum_{t=1}^T (Z_t-M^*)^2 \preceq \delta (M')^2 + \sum_{t=1}^T (Z_t-M')^2. \end{equation*}
\end{lemma}
\begin{proof}
Write $M' = M^* + D$. Then we have
\begin{dgroup*}
\begin{dmath*} \delta(M')^2 + \sum_{t=1}^T (Z_t - M')^2
\end{dmath*}\begin{dmath*}  = \delta(M^*+D)^2 + \sum_{t=1}^T (Z_t - M^* - D)^2 
\end{dmath*}\begin{dmath*}  = \delta(M^*)^2 + \sum_{t=1}^T (Z_t - M^*)^2 + \left[\delta M^* + \sum_{t=1}^T M^*-Z_t\right]D + D\left[\delta M^* + \sum_{t=1}^T M^*-Z_t\right] + TD^2 
\end{dmath*}\begin{dmath*}  = \delta(M^*)^2 + \sum_{t=1}^T (Z_t-M^*)^2 + TD^2 
\end{dmath*}\begin{dmath*}  \succeq \delta(M^*)^2 + \sum_{t=1}^T (Z_t-M^*)^2,
\end{dmath*}
\end{dgroup*}
which completes the lemma.
\end{proof}
Setting $\delta$ to $0$, we now see that $\frac{1}{T} \sum_{t=1}^T Z_t$ is the 
optimal value of $M^*$ for any $W \succeq 0$, in analogy with the vector case. We 
remark that Lemma~\ref{lem:global-opt} is almost purely algebraic, and only relies 
on the property that $D^2 \succeq 0$ for any symmetric matrix $D$. We could obtain 
a similar lemma for any partially ordered algebra equipped with an adjoint operator 
$*$ such that $a^{*}a \succeq 0$ for all elements $a$ of the algebra.

@@

\begin{lemma}
\label{lem:ftrl-k}
Suppose that for all $1 \leq t \leq T+1$, there 
exists a global minimizer $M_t$ of 
$\psi(M) + \sum_{s=1}^{t-1} f_s(M)$. Then for all $M$,
\begin{dmath} \sum_{t=1}^T f_t(M_t)-f_t(M) \leq_{\sK} \psi(M) - \psi(M_1) + \sum_{t=1}^T f_t(M_{t}) - f_t(M_{t+1}). \end{dmath}
\end{lemma}
\begin{proof}
We will prove the lemma by induction on $T$. Note that the lemma is equivalent to showing that 
\[ \psi(M_1) + \sum_{t=1}^T f_t(M_{t+1}) \leq_{\sK} \psi(M) + \sum_{t=1}^T f_t(M) \]
for all $M$.

In the base case $T = 0$, we have
\[ \psi(M_1) \leq_{\sK} \psi(M), \]
which follows from the fact that $M_1$ is a global minimizer of $\psi$ and 
hence $\psi(M_1) \leq_{\sK} \psi(M)$ for all $M$. For the inductive step, suppose that
\[ \sum_{t=1}^{T-1} f_t(M_{t+1}) \leq_{\sK} \psi(M) \sum_{t=1}^{T-1} f_t(M) \]
for all $M$, and invoke this for the particular choice $M = M_{T+1}$. Then we have
\begin{align*}
\lefteqn{\psi(M_1) + \sum_{t=1}^T f_t(M_{t+1})} \\
 &= \psi(M_1) + \sum_{t=1}^{T-1} f_t(M_{t+1}) + f_T(M_{T+1}) \\
 &\leq_{\sK} \psi(M_{T+1}) + \sum_{t=1}^{T-1} f_t(M_{T+1}) + f_T(M_{T+1}) \\
 &= \psi(M_{T+1}) + \sum_{t=1}^T f_t(M_{T+1}) \\
 &\leq_{\sK} \psi(M) + \sum_{t=1}^T f_t(M)
\end{align*}
for all $M$, where we use the fact that $M_{T+1}$ is a global minimizer of $\psi(M) + \sum_{t=1}^T f_t(M)$ for 
the last inequality. This completes the inductive hypothesis and hence the proof.
\end{proof}

As a corollary, we have (compare to Proposition~\ref{prop:ftrl-aux}):
\begin{corollary}
\label{cor:ftrl-k-aux}
Suppose that we choose $M_t = \frac{1}{t} \sum_{s=1}^{t-1} Z_s$. Then
\[ \sum_{t=1}^T (Z_t - M_t)^2 \preceq 2\sum_{t=1}^T (Z_t - M^*)^2 + 6I, \]
for $M^* \eqdef \frac{1}{T} \sum_{t=1}^T Z_t$.
\end{corollary}
\begin{proof}
The key observation is the \emph{matrix Young's inequality}: 
$AB + BA \preceq \frac{1}{\gamma}A^2 + \gamma B^2$ for all symmetric $A$, $B$ and 
all $\gamma > 0$. (This follows immediately upon expanding 
$(A/\sqrt{\gamma} - \sqrt{\gamma} B)^2 \succeq 0$.) We then note that 
$M_t$ obeys Lemma~\ref{lem:ftrl-k} with $\psi(M) = M^2$ and $f_t(M) = (M-Z_t)^2$. 
Hence we have
\begin{dgroup*}
\begin{dmath*} \sum_{t=1}^T (Z_t - M_t)^2 - (Z_t - M^*)^2
\end{dmath*}\begin{dmath*} = (M^*)^2 + \sum_{t=1}^T (Z_t - M_t)^2 - (Z_t - M_{t+1})^2 
\end{dmath*}\begin{dmath*} = (M^*)^2 + \sum_{t=1}^T Z_t(M_{t+1} - M_t) + (M_{t+1}-M_t)Z_t + M_t^2 - M_{t+1}^2 
\end{dmath*}\begin{dmath*} = (M^*)^2 + M_1^2 - M_{T+1}^2 + \sum_{t=1}^T \frac{1}{t+1}\left[ Z_t(Z_t-M_t) + (Z_t-M_t)Z_t\right] 
\end{dmath*}\begin{dmath*} \leq I + \sum_{t=1}^T \frac{Z_t^2}{\gamma (t+1)^2} + \gamma (Z_t - M_t)^2 
\end{dmath*}\begin{dmath*} \leq I + \frac{I}{\gamma} + \gamma \sum_{t=1}^T (Z_t - M_t)^2.
\end{dmath*}
\end{dgroup*}
Re-arranging yields
\begin{equation*}
\sum_{t=1}^T (Z_t - M_t)^2 \leq \frac{1}{1-\gamma} \left( \frac{1+\gamma}{\gamma}I + \sum_{t=1}^T (Z_t - M^*)^2\right),
\end{equation*}
which for $\gamma = \frac{1}{2}$ again yields the desired result.
\end{proof}
Combining Proposition~\ref{prop:maeg} with Corollary~\ref{cor:ftrl-k-aux} gives a 
regret bound for the MEG-Variance algorithm, whose updates are given in 
Table~\ref{tab:algorithms}.
\begin{corollary}
\label{cor:maeg-variance}
The MEG-Variance algorithm achieves a regret bound of
\begin{dmath}\Regret(w) \leq \frac{\log(n)}{\eta} + \eta \left[ \sum_{t=1}^T \Tr(W(Z_t-M^*)^2) + 4\log(T+1) \right].\end{dmath} 
\end{corollary}
